---
title: "GenomeProt Report"
date: "`r format(Sys.Date())`"
output: html_document
layout: single
params:
  directory:
    value: x
  file:
    value: x
---

```{r setup, include = FALSE, warning = FALSE}
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(readr)
  library(knitr)
  library(scales)
  library(openxlsx)
})

options(scipen = 999)

wb <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb, "ORF_by_location")
openxlsx::addWorksheet(wb, "ORF_by_tx_biotype")

df <- readr::read_tsv(file.path(params$directory, params$file))
df_var_peptides <- df %>% dplyr::filter(grepl("_var", accession))

bad_pep <- df %>% dplyr::filter(is.na(transcript_id))
df <- df %>% dplyr::filter(!is.na(transcript_id))
```

##### Definitions:
ORF: Open reading frame.

Canonical ORF: ORF annotated in UniProt or RefSeq databases.

Noncanonical ORF (ncORF): ORF not annotated in UniProt or RefSeq databases.

UMP (Uniquely Mapped Peptide): Peptide uniquely mapped to a single ORF or protein.

Peptide evidence: Feature supported by any mapped peptide.

Transcript-specific or gene-specific peptide: Peptide uniquely mapped to a specific transcript or gene.

Peptide-confirmed: Feature distinguished by transcript-specific or gene-specific peptides.

### Summary of peptides, transcripts and ORFs

```{r, eval = TRUE, echo = FALSE}
peptotal <- length(base::unique(df$peptide))
uniqpeptides <- nrow(df %>%
                     dplyr::group_by(peptide) %>%
                     dplyr::filter(length(base::unique(accession)) == 1) %>%
                     dplyr::slice_head(n = 1))

txtotal <- length(base::unique(df$transcript_id))
txuniqtotal <- nrow(df %>%
                    dplyr::filter(transcript_identified == TRUE) %>%
                    dplyr::select("transcript_id") %>%
                    base::unique())

genetotal <- length(base::unique(df$gene_id))
geneuniqtotal <- df %>%
                 dplyr::filter(gene_identified == TRUE) %>%
                 dplyr::select("gene_id") %>%
                 base::unique() %>%
                 NROW()

orftotal <- base::unique(df$accession) %>% NROW()
orfuniqtotal <- df %>%
                dplyr::filter(orf_identified == TRUE) %>%
                dplyr::select(accession) %>%
                base::unique() %>%
                NROW()
```

Total Peptides: `r paste0(peptotal)`

UMPs: `r paste0(uniqpeptides)`

Transcripts with peptide evidence: `r paste0(txtotal)`

Peptide-confirmed transcripts: `r paste0(txuniqtotal)`

Genes with peptide evidence: `r paste0(genetotal)`

Peptide-confirmed genes: `r paste0(geneuniqtotal)`

Proteins/ORFs with peptide evidence: `r paste0(orftotal)`

Proteins/ORFs with UMPs: `r paste0(orfuniqtotal)`

### Summary of peptides mapped to contaminants, decoys or that weren't mapped to genomic loci

```{r, eval = TRUE, echo = FALSE}
bad_rev <- bad_pep %>% dplyr::filter(startsWith(PID, "rev"))
if (nrow(bad_rev) == 0) {
  bad_rev_num <- 0
} else {
  bad_rev_num <- length(base::unique(bad_rev$peptide))
}

bad_contam <- bad_pep %>% dplyr::filter(startsWith(PID, "sp"))
if (nrow(bad_contam) == 0) {
  bad_contam_num <- 0
} else {
  bad_contam_num <- length(base::unique(bad_contam$peptide))
}

bad_multi_loc <- bad_pep %>% dplyr::filter(!startsWith(PID, "sp") & !startsWith(PID, "rev"))
if (nrow(bad_multi_loc) == 0) {
  bad_multi_loc_num <- 0
} else {
  bad_multi_loc_num <- length(base::unique(bad_multi_loc$peptide))
}
```

Total peptides mapped to decoys: `r paste0(bad_rev_num)`

Total peptides mapped to contaminants: `r paste0(bad_contam_num)`

Total other peptides that weren't mapped (likely belong to ORFs with multiple genomic locations): `r paste0(bad_multi_loc_num)`

### Peptide-level data

```{r, eval = TRUE, echo = FALSE}
# total uniquely mapped peptides
a <- df %>%
     dplyr::filter(peptide_ids_orf == TRUE) %>%
     dplyr::select(peptide) %>%
     base::unique() %>%
     NROW()

# Transcript-specific peptides
b <- df %>%
     dplyr::filter(peptide_ids_transcript == TRUE) %>%
     dplyr::select(peptide) %>%
     base::unique() %>%
     NROW()

# Gene-specific peptides
c <- df %>%
     dplyr::filter(peptide_ids_gene == TRUE) %>%
     dplyr::select(peptide) %>%
     base::unique() %>%
     NROW()

# Peptides spanning splice junctions
d <- df %>%
     dplyr::filter(number_of_mapped_exons > 1) %>%
     dplyr::select(peptide) %>%
     base::unique() %>%
     NROW()

# Peptides mapped to multiple ORFs from multiple genes
e <- nrow(df %>%
          dplyr::group_by(peptide) %>%
          dplyr::filter(length(base::unique(PID)) > 1 & length(base::unique(gene_id)) > 1) %>%
          dplyr::slice_head(n = 1))

# Make stats
item <- c("Uniquely mapped peptides",
          "Transcript-specific peptides",
          "Gene-specific peptides",
          "Peptides spanning splice junctions",
          "Peptides mapped to multiple ORFs from multiple genes")

total <- c(a, b, c, d, e)
stats <- data.frame(item, total)
knitr::kable(stats, format = "html", col.names = c(" ", ""))
```

### Transcript-level data

```{r, eval = TRUE, echo = FALSE}
# Known transcripts with peptide evidence
f <- df %>%
     dplyr::filter(!startsWith(transcript_id, "Bambu")) %>%
     dplyr::select(transcript_id) %>%
     base::unique() %>%
     NROW()

# Novel transcripts with peptide evidence
g <- df %>%
     dplyr::filter(startsWith(transcript_id, "Bambu")) %>%
     dplyr::select(transcript_id) %>%
     base::unique() %>%
     NROW()

# Known transcripts encoding ORFs with UMPs
h <- df %>%
     dplyr::filter(orf_identified == TRUE) %>%
     dplyr::filter(!startsWith(transcript_id, "Bambu")) %>%
     dplyr::select(transcript_id) %>%
     base::unique() %>%
     NROW()

# Novel transcripts encoding ORFs with UMPs
i <- df %>%
     dplyr::filter(orf_identified == TRUE) %>%
     dplyr::filter(startsWith(transcript_id, "Bambu")) %>%
     dplyr::select(transcript_id) %>%
     base::unique() %>%
     NROW()

# Peptide-confirmed known transcripts
j <- df %>%
     dplyr::filter(transcript_identified == TRUE & !startsWith(transcript_id, "Bambu")) %>%
     dplyr::select(transcript_id) %>%
     base::unique() %>%
     NROW()

# Peptide-confirmed novel transcripts
k <- df %>%
     dplyr::filter(transcript_identified == TRUE & startsWith(transcript_id, "Bambu")) %>%
     dplyr::select(transcript_id) %>%
     base::unique() %>%
     NROW()

# Peptide-confirmed novel transcripts encoding annotated ORFs with UMPs
l <- df %>%
     dplyr::filter(transcript_identified == TRUE & startsWith(transcript_id, "Bambu")) %>%
     dplyr::filter(!startsWith(accession, "ORF_")) %>%
     dplyr::select(accession) %>%
     base::unique() %>%
     NROW()

# Peptide-confirmed novel transcripts encoding unannotated ORFs with UMPs
m <- df %>%
     dplyr::filter(transcript_identified == TRUE & startsWith(transcript_id, "Bambu")) %>%
     dplyr::filter(startsWith(accession, "ORF_")) %>%
     dplyr::select(accession) %>%
     base::unique() %>%
     NROW()

# Make stats
item <- c("Annotated (known) transcripts with peptide evidence",
          "Unannotated (novel) transcripts with peptide evidence",
          "Annotated transcripts encoding ORFs with UMPs",
          "Unannotated transcripts encoding ORFs with UMPs",
          "Peptide-confirmed annotated transcripts",
          "Peptide-confirmed unannotated transcripts",
          "Peptide-confirmed unannotated transcripts encoding canonical ORFs with UMPs",
          "Peptide-confirmed unannotated transcripts encoding noncanonical ORFs with UMPs")

total <- c(f, g, h, i, j, k, l, m)
stats <- data.frame(item, total)
knitr::kable(stats, format = "html", col.names = c(" ", ""))
```

```{r variant_protein_isoform, echo = FALSE}
if (nrow(df_var_peptides) > 0) {
  cat("### Variant protein isoform data\n\n")

  # Number of variant ORFs with UMPs
  var_p <- df_var_peptides %>%
           dplyr::filter(peptide_ids_orf == TRUE) %>%
           dplyr::select(accession) %>%
           base::unique() %>%
           NROW()

  # Number of canonical variant ORFs with UMPs
  var_p_canonical <- df_var_peptides %>%
                     dplyr::filter(peptide_ids_orf == TRUE) %>%
                     dplyr::select(accession) %>%
                     dplyr::filter(!startsWith(accession, "ORF_")) %>%
                     base::unique() %>%
                     NROW()

  # Number of canonical variant ORFs with UMPs
  var_p_novel <- df_var_peptides %>%
                 dplyr::filter(peptide_ids_orf == TRUE) %>%
                 dplyr::select(accession) %>%
                 dplyr::filter(startsWith(accession, "ORF_")) %>%
                 base::unique() %>%
                 NROW()

  # Number of variant protein specific UMPs
  var_pep <- df_var_peptides %>%
             dplyr::filter(peptide_ids_orf == TRUE) %>%
             dplyr::select(peptide) %>%
             base::unique() %>%
             NROW()

  # Number of canonical variant protein specific UMPs
  var_pep_novel <- df_var_peptides %>%
                   dplyr::filter(peptide_ids_orf == TRUE) %>%
                   dplyr::select(accession, peptide) %>%
                   dplyr::filter(startsWith(accession, "ORF_")) %>%
                   dplyr::select(peptide) %>%
                   base::unique() %>%
                   NROW()

  # Number of canonical variant protein specific UMPs
  var_pep_known <- df_var_peptides %>%
                   dplyr::filter(peptide_ids_orf == TRUE) %>%
                   dplyr::select(accession, peptide) %>%
                   dplyr::filter(!startsWith(accession, "ORF_")) %>%
                   dplyr::select(peptide) %>%
                   base::unique() %>%
                   NROW()

  item_var <- c("Variant proteins with UMPs",
                "Canonical variant proteins with UMPs",
                "Noncanonical variant proteins with UMPs",
                "Variant UMPs",
                "Canonical variant protein derived UMPs",
                "Noncanonical variant protein derived UMPs")

  total_var <- c(var_p, var_p_canonical, var_p_novel, var_pep, var_pep_known, var_pep_novel)
  stats_variant_proteins <- data.frame(item_var, total_var)
  knitr::kable(stats_variant_proteins, format = "html", col.names = c(" ", ""))
}
```

### Protein/ORF-level data

```{r, eval = TRUE, echo = FALSE}
# Annotated ORFs with peptide evidence
n <- df %>%
     dplyr::filter(!startsWith(accession, "ORF_")) %>%
     dplyr::select(accession) %>%
     base::unique() %>%
     NROW()

# Unannotated ORFs with peptide evidence
o <- df %>%
     dplyr::filter(startsWith(accession, "ORF_")) %>%
     dplyr::select(accession) %>%
     base::unique() %>%
     NROW()

# Annotated ORFs with UMPs
p <- df %>%
     dplyr::filter(orf_identified == TRUE & !startsWith(accession, "ORF_")) %>%
     dplyr::select(accession) %>%
     base::unique() %>%
     NROW()

# Unannotated ORFs with UMPs
q <- df %>%
     dplyr::filter(orf_identified == TRUE & startsWith(accession, "ORF_")) %>%
     dplyr::select(accession) %>%
     base::unique() %>%
     NROW()

# Make stats
item <- c("Canonical ORFs with peptide evidence",
          "Noncanonical ORFs with peptide evidence",
          "Canonical ORFs with UMPs",
          "Noncanonical ORF with UMPs")

total <- c(n, o, p, q)
stats <- data.frame(item, total)
knitr::kable(stats, format = "html", col.names = c(" ", ""))
```

### Figures

#### Peptides by mapping status (Total peptides: `r peptotal`)

```{r, eval = TRUE, echo = FALSE}
plot1 <- data.frame(Category = c("Shared peptides", "Uniquely mapped peptides"),
                    count = c(peptotal - a, a))
plot1$percent <- round(100 * plot1$count / sum(plot1$count), 1)
plot1$label <- paste0(plot1$count, "(", plot1$percent, "%)")

pepplot <- ggplot2::ggplot(plot1, ggplot2::aes(x = "", y = count, fill = Category)) +
           ggplot2::geom_col(width = 1, color = "white") +
           ggplot2::coord_polar(theta = "y") +
           ggplot2::geom_text(ggplot2::aes(label = label),
                              position = ggplot2::position_stack(vjust = 0.5), size = 5) +
           ggplot2::theme_void() +
           ggplot2::labs(fill = "Category") +
           ggplot2::scale_fill_manual(values = c("Shared peptides" = "#D3D3D3", "Uniquely mapped peptides" = "lightgreen"))

suppressWarnings(print(pepplot))
```

#### Number of UMPs per ORF (shown for >10 ORFs)

```{r, eval = TRUE, echo = FALSE}
umps_per_protein <- df %>%
                    dplyr::filter(orf_identified == TRUE) %>%
                    dplyr::group_by(accession) %>%
                    dplyr::summarise(number_ump = length(base::unique(peptide))) %>%
                    dplyr::ungroup() %>%
                    dplyr::mutate(orf_type = dplyr::case_when(
                      startsWith(accession, "ORF") ~ "unannotated",
                      TRUE ~ "annotated"
                    ))

tallied <- umps_per_protein %>%
           dplyr::group_by(orf_type, number_ump) %>%
           dplyr::summarise(n = dplyr::n()) %>%
           dplyr::filter(n > 10) %>%
           dplyr::rename("ORF type" = orf_type)

ump_per_protein_plot <- ggplot2::ggplot(tallied) +
                        ggplot2::geom_bar(ggplot2::aes(x = number_ump, y = n, fill = `ORF type`), position = "stack", stat = "identity") +
                        ggplot2::ylab("Number of ORFs") +
                        ggplot2::xlab("Number of UMPs per ORF") +
                        ggplot2::scale_x_continuous(breaks = scales::breaks_pretty()) +
                        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
                        ggplot2::scale_fill_manual(values = c("unannotated" = "deepskyblue3", "annotated" = "darkgreen")) +
                        ggplot2::theme_bw()

suppressWarnings(print(ump_per_protein_plot))
```

#### Peptides by number of exons spanned

```{r, eval = TRUE, echo = FALSE}
summary_df <- df %>%
              dplyr::group_by(peptide) %>%
              dplyr::slice_head(n = 1) %>%
              dplyr::ungroup() %>%
              dplyr::group_by(number_of_mapped_exons) %>%
              dplyr::summarise(total = dplyr::n())

exons_plot <- ggplot2::ggplot(summary_df) +
              ggplot2::geom_bar(ggplot2::aes(x = number_of_mapped_exons, y = total, fill = number_of_mapped_exons), stat = "identity") +
              ggplot2::ylab("Number of peptides") +
              ggplot2::xlab("Number of exons spanned") +
              ggplot2::theme_bw() +
              ggplot2::theme(legend.position = "none")

suppressWarnings(print(exons_plot))
```

#### Noncanonical ORFs with UMPs by localisation

```{r, eval = TRUE, echo = FALSE}
summary_df <- df %>%
              dplyr::filter(peptide_ids_orf == TRUE & startsWith(accession, "ORF")) %>%
              dplyr::select(accession, localisation) %>%
              base::unique() %>%
              dplyr::group_by(localisation) %>%
              dplyr::summarise(Count = dplyr::n())

openxlsx::writeData(wb, sheet = "ORF_by_location", x = summary_df)

plot_orf_local <- ggplot2::ggplot(summary_df) +
                  ggplot2::geom_bar(ggplot2::aes(x = localisation, y = Count, fill = localisation), stat = "identity") +
                  ggplot2::ylab("Number of ncORFs") +
                  ggplot2::xlab("Localisation") +
                  ggplot2::scale_fill_brewer(palette = "Dark2") +
                  ggplot2::theme_bw() +
                  ggplot2::theme(legend.position = "none")

suppressWarnings(print(plot_orf_local))
```

#### Noncanonical ORFs with UMPs by simplified transcript biotype

```{r, eval = TRUE, echo = FALSE}
summary_df <- df %>%
              dplyr::filter(orf_identified == TRUE & startsWith(accession, "ORF")) %>%
              dplyr::select(accession, simplified_biotype) %>%
              base::unique() %>%
              dplyr::group_by(simplified_biotype) %>%
              dplyr::summarise(Count = dplyr::n())

openxlsx::writeData(wb, sheet = "ORF_by_tx_biotype", x = summary_df)

plot_orf_biotype <- ggplot2::ggplot(summary_df) +
                    ggplot2::geom_bar(ggplot2::aes(x = simplified_biotype, y = Count, fill = simplified_biotype), stat = "identity") +
                    ggplot2::ylab("Number of ncORFs") +
                    ggplot2::xlab("Transcript biotype") +
                    ggplot2::scale_fill_brewer(palette = "Set3") +
                    ggplot2::theme_bw() +
                    ggplot2::theme(legend.position = "none",
                                   axis.text.x = ggplot2::element_text(angle = -45, hjust = 0, vjust = 1))

suppressWarnings(print(plot_orf_biotype))
```

#### Lengths of annotated and unannotated ORFs (with UMPs)

```{r, eval = TRUE, echo = FALSE}
summary_df <- df %>%
              dplyr::filter(orf_identified == TRUE) %>%
              dplyr::select(accession, protein_length, orf_type) %>%
              base::unique() %>%
              dplyr::rename("ORF type" = orf_type)

plot_orf_lengths <- ggplot2::ggplot(summary_df) +
                    ggplot2::geom_density(ggplot2::aes(x = protein_length, y = ..scaled.., color = `ORF type`)) +
                    ggplot2::ylab("Scaled density") +
                    ggplot2::xlab("ORF Length (0-1000 AA)") +
                    ggplot2::scale_color_manual(values = c("annotated" = "darkgreen", "unannotated" = "deepskyblue3")) +
                    ggplot2::theme_bw() +
                    ggplot2::xlim(0, 1000)

suppressWarnings(print(plot_orf_lengths))
```

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
# save stats
openxlsx::saveWorkbook(wb, file.path(params$directory, "ncORF_stats.xlsx"), overwrite = TRUE)

# save all PDFs
pdf(file.path(params$directory, "report_images", "peptides-by-mapping-status.pdf"), w = 6, h = 4)
suppressWarnings(print(pepplot))
invisible(dev.off())

pdf(file.path(params$directory, "report_images", "umps-per-orf.pdf"), w = 6, h = 4)
suppressWarnings(print(ump_per_protein_plot))
invisible(dev.off())

pdf(file.path(params$directory, "report_images", "exons-spanned-by-peptides.pdf"), w = 6, h = 4)
suppressWarnings(print(exons_plot))
invisible(dev.off())

pdf(file.path(params$directory, "report_images", "unannotated-orf-localisations.pdf"), w = 6, h = 4)
suppressWarnings(print(plot_orf_local))
invisible(dev.off())

pdf(file.path(params$directory, "report_images", "unannotated-orf-tx-biotypes.pdf"), w = 6, h = 4)
suppressWarnings(print(plot_orf_biotype))
invisible(dev.off())

pdf(file.path(params$directory, "report_images", "lengths-annotated-v-unannotated-orfs.pdf"), w = 6, h = 4)
suppressWarnings(print(plot_orf_lengths))
invisible(dev.off())
```